# Agglomeration {#sec-agglomeration}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

```{r, message=FALSE}
library(mia)
data("GlobalPatterns", package = "mia")
tse <- GlobalPatterns
```

In this chapter, we discuss agglomeration, which involves summing data within
specific groups. For example, we can agglomerate data to the phylum taxonomy
level. This process begins by identifying which phyla are present in the data.
Subsequently, we group the data according to these phyla and aggregate the
counts. The resulting dataset will have features corresponding to each phylum,
with counts aggregated from the lower-level taxa associated with them.

## Agglomerate data to certain rank {#sec-data-agglomeration}

One of the main applications of taxonomic information in regards to count data
is to agglomerate count data on taxonomic levels and track the influence of
changing conditions through these levels. For this `mia` contains the
`agglomerateByRank()` function. The ideal location to store the agglomerated data
is as an alternative experiment(see [@sec-alt-exp]).

```{r}
# Transform data
tse <- transformAssay(tse, assay.type = "counts", method = "relabundance")
# Agglomerate
altExp(tse, "Family") <- agglomerateByRank(
    tse, rank = "Family", agglomerate.tree = TRUE)
altExp(tse, "Family")
```

If multiple assays (counts and relabundance) exist, both will be agglomerated.

```{r}
assayNames(tse)
assayNames(altExp(tse, "Family"))
```

```{r}
assay(altExp(tse, "Family"), "relabundance")[1:5, 1:7]
```

```{r taxinfo_altexp_example}
assay(altExp(tse, "Family"), "counts")[1:5, 1:7]
```

`altExpNames` now consists of `Family` level data. This can be extended to use
any taxonomic level listed in `taxonomyRanks(tse)`.

We can also aggregate the data across all available ranks in one step using
`agglomerateByRanks()`. The function returns `TreeSE` including agglomerated
objects in `altExp` slot.

```{r}
#| label: agglomerateranks

tse <- agglomerateByRanks(tse)
altExpNames(tse)
```

### Total abundance of certain taxa

When total abundances of certain phyla are of relevance, the data is initially
agglomerated by Phylum. Then, similar steps as in the case of non-agglomerated
data are followed.

```{r}
# Get the agglomerated data from altExp
tse_phylum <- altExp(tse, "Phylum")

# Subset by feature and remove NAs
tse_sub <- tse_phylum[
    rowData(tse_phylum)$Phylum %in% c("Actinobacteria", "Chlamydiae")
        &!is.na(rowData(tse_phylum)$Phylum), ]

# Show dimensions
dim(tse_sub)
```

::: {.callout-note}
## Note
As data was agglomerated, the number of rows should equal the
number of phyla used to index (in this case, just 2).
:::

Alternatively:

```{r}
# Store features of interest into phyla
phyla <- c("Actinobacteria", "Chlamydiae")
# subset by feature
tse_sub <- tse_phylum[phyla, ]
# Show dimensions
dim(tse_sub)
```

## Agglomerate based on prevalence

Rare taxa can also be aggregated into a single group "Other" instead of 
filtering them out. A suitable function for this is `agglomerateByPrevalence()`.
The number of rare taxa is higher on the species level, which causes the need 
for data agglomeration by prevalence.

```{r}
altExp(tse, "Species_byPrevalence") <- agglomerateByPrevalence(
    tse,
    rank = "Species",
    other.label = "Other",
    prevalence = 5 / 100,
    detection = 1 / 100,
    as.relative = TRUE)
altExp(tse, "Species_byPrevalence")

assay(altExp(tse, "Species_byPrevalence"), "relabundance")[1:5, 1:3]
```

## Aggregate data based on variable

`agglomerateByRank()` aggregates the data taking into account the taxonomy
information. For more flexible aggregations, there is available method
`agglomerateByVariable()`. For instance, we can aggregate the data by sample
types.

```{r}
#| label: aggregate_samples

# Agglomerate samples based on type
tse_sub <- agglomerateByVariable(tse, by = "cols", f = "SampleType")
tse_sub
```

[@sec-taxa-clustering] introduces how cluster information can be utilized to
agglomerate data.

## Subset based on prevalence

In addition to agglomeration, we can subset the data based on prevalence.
Using `subsetByPrevalent()`, we can filter for taxa that exceed a specified
prevalence threshold. Alternatively, `subsetByRare()` allows us to filter for taxa
that do not exceed the threshold.

```{r}
#| label: subset_by_rare

tse_sub <- subsetByRare(tse, rank = "Genus", detection = 0.01, prevalence = 0.1)
tse_sub
```

