# Community Typing {#sec-clustering}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

Community typing in microbial ecology involves identifying distinct microbial
communities by recognizing patterns in the data. The community types are
described based on taxa characteristics, representing each community.
The samples are described
by community type assignments, which are defined by the ecosystem features in
the sample. Community typing techniques can typically be divided
into unsupervised
clustering and dimensionality reduction techniques, where clustering is more
commonly in use.

In this chapter, we first walk you through some common clustering schemes in use
for microbial community ecology, and then focus in more detailed community
typing techniques based on data dimensionality reduction.

## Clustering

Clustering techniques aim to find groups, called clusters, that share a pattern
in the data. In the microbiome context, clustering techniques are included in
microbiome community typing methods. For example, clustering allow samples to be
distinguished from each other based on their microbiome community composition.
Clustering scheme consists of two steps, the first is to compute the sample
dissimilarities with a given distance metrics, and the second is to form the
clusters based on the dissimilarity matrix. The data can be clustered either
based on features or samples. The examples below
are focused on sample clustering.

There are multiple clustering algorithms available. `bluster` is a 
Bioconductor package providing tools for clustering data in the 
`SummarizedExperiment` container. It offers multiple algorithms such 
as hierarchical clustering, DBSCAN, and K-means. 

```{r load_bluster}
# Load dependencies
library(bluster)
```

In the examples of this chapter we use *peerj13075* data for
microbiota community
typing. This chapter illustrates how different results can be obtained depending
on the choice of the algorithm. To reduce calculation time we decided to
agglomerate taxa onto 'Class' level and filter out the least prevalent taxa as
well as less commonly detected within a sample resulting in 25 features in the
data.

```{r load-pkg-data1}
library(mia)
data("peerj13075", package = "mia")
tse <- peerj13075

# Filter out most taxa to ease the calculation
altExp(tse, "prevalent") <- agglomerateByPrevalence(
    tse, rank = "class", prevalence = 20/100, detection = 1/100)
```

### Hierarchical clustering

The hierarchical clustering aims to find hierarchy between
samples/features. There are two approaches: agglomerative ("bottom-up")
and divisive ("top-down"). In agglomerative approach, each observation is first
in a unique cluster, after which the algorithm continues to agglomerate similar
data points into clusters. The divisive approach, instead, starts with one
cluster that contains all observations. Clusters are split recursively into
clusters that differ the most. The clustering can be continued until each
cluster contains only one observation.

In this example we use `addCluster()` function from `mia` to  cluster the data. 
`addCluster()` function allows to choose a clustering algorithm and offers multiple 
parameters to shape the result. `HclustParam()` parameter is chosen for hierarchical 
clustering. `HclustParam()` parameter itself has parameters on its own 
[HclustParam documentation](https://rdrr.io/github/LTLA/bluster/man/HclustParam-class.html).
A parameter, `by` defines whether to cluster features or samples. Here we 
cluster counts data, for which we compute the dissimilarities with the 
Bray-Curtis distance. Here, again, we use `ward.d2` method.  Returning 
statistical information can be chosen by using 
`full` parameter. Finally, the `clust.col` parameter allows us to choose the 
name of the column in the colData (default name is `clusters`).

```{r hclust1}
library(vegan)

set.seed(174923)
altExp(tse, "prevalent") <- addCluster(
    altExp(tse, "prevalent"),
    assay.type = "counts",
    by = "cols",
    HclustParam(method = "ward.D2", dist.fun = getDissimilarity, metric = "bray"),
    full = TRUE,
    clust.col = "hclust")

table(altExp(tse, "prevalent")$hclust)
```

Hierarchical clustering suggests six clusters to describe the data.

Now, we visualize the hierarchical structure of the clusters with a dendrogram
tree. In dendrograms, the tree is split where the branch length is the largest.
In each splitting point, the tree is divided into two clusters leading to the
hierarchy. In this example, each sample is labelled by their dominant taxon
to visualize ecological differences between the clusters.

```{r hclust2}
library(dendextend)

# Get hclust data from metadata
hclust_data <- metadata(altExp(tse, "prevalent"))$clusters$hclust

# Get the dendrogram object
dendro <- as.dendrogram(hclust_data)
k <- length(unique(altExp(tse, "prevalent")$hclust))

# Color the branches by clusters
cols <- scales::hue_pal()(k)

# Order the colors to be equivalent to the factor levels
cols <- cols[c(1,4,5,3,2,6)]
dend <- color_branches(dendro, k = k, col = unlist(cols))

# Label the samples by their dominant taxon
altExp(tse, "prevalent") <- addDominant(altExp(tse, "prevalent"))
labels(dend) <- altExp(tse, "prevalent")$dominant_taxa
dend <- color_labels(dend, k = k, col = unlist(cols))

# Adjust frame for the figure
par(mar=c(9, 3, 0.5, 0.5))
# Plot dendrogram
dend |> set("labels_cex", 0.8) |> plot()
```

We can also visualize the clusters by projecting the data onto
two-dimensional
surface that captures the most variability in the data. In this example, we use
multi-dimensional scaling (MDS) (see [@sec-unsupervised-ordinantion]).

```{r hclust3}
library(scater)

# Add the MDS dimensions for plotting
altExp(tse, "prevalent") <- runMDS(
    altExp(tse, "prevalent"),
    assay.type = "counts",
    FUN = vegan::vegdist,
    method = "bray")
```

```{r hclust4, fig.width=6, fig.height=4}
plotReducedDim(
    altExp(tse, "prevalent"), "MDS", colour_by = "hclust",
    theme_size = 13, point_size = 4) +
    labs(color = "HClust", title = "Hclust") + theme_minimal()
```

### Agglomeration based on clusters {#sec-taxa-clustering}

Another way to agglomerate the data is to cluster the taxa. To do so,
we usually start by doing a compositionality aware transformation such as CLR,
followed by the application of a standard clustering method.

Here is an example that does a CLR transformation followed by the hierarchical
clustering algorithm. We will use the kendall dissimilarity.

```{r taxa_clustering}
# The result of the CLR transform is stored in the assay clr
tse <- transformAssay(tse, method = "clr", pseudocount = 1)

tse <- transformAssay(
    tse, assay.type = "clr", method = "standardize", MARGIN = "rows")

# Declare the Kendall dissimilarity computation function
kendall_dissimilarity <- function(x) {
    as.dist(1 - cor(t(x), method = "kendall"))
}

# Cluster (with Kendall dissimilarity) on the features of the `standardize`
# assay
tse <- addCluster(
    tse,
    assay.type = "standardize",
    clust.col = "hclustKendall",
    MARGIN = "rows",
    HclustParam(dist.fun = kendall_dissimilarity, method = "ward.D2"))
```

Results are stored in the `rowData` column specified
with the `clust.col` parameter.

To better visualize the results and the distribution of the clusters, we can
plot the histogram of the clusters.

```{r taxa_clustering_histogram}
library(ggplot2)

p <- ggplot(rowData(tse), aes(x = hclustKendall)) +
    geom_bar() +
    labs(title = "CAG size distribution (1 - tau)",
         x = "Clusters", y = "Feature count (n)")
p
```

Then we can merge data to clusters.

```{r taxa_clustering_row_merge}
# Aggregate clusters as a sum of each cluster values
tse_merged <- agglomerateByVariable(tse, by = "rows", f = "hclustKendall")
tse_merged
```

We can note that it worked as planned since there are now as many rows as there
were clusters.

You can find more information on agglomeration from [@sec-agglomeration].

### Dirichlet Multinomial Mixtures (DMM)

This section focuses on
[Dirichlet-Multinomial MixturenModel](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0030126)
analysis. It is a Bayesian clustering technique that allows to search for sample
patterns that reflect sample similarity in the data using both prior information
and the data.

In this example, we cluster the data with DMM clustering. In the example below,
we calculate model fit using Laplace approximation ro reduce the
calculation time.
The cluster information is added in the metadata with an
optional `name`. In this
example, we use a prior assumption that the optimal number of clusters to
describe the data would be in the range from 1 to 6.

```{r dmm1}
# Run the model and calculates the most likely number of clusters from 1 to 6
set.seed(1495723)
altExp(tse, "prevalent") <- addCluster(
    altExp(tse, "prevalent"),
    assay.type = "counts",
    name = "DMM",
    DmmParam(k = 1:6, type = "laplace"),
    MARGIN = "samples",
    full = TRUE,
    clust.col = "dmmclust")
```

The plot below represents the Laplace approximation to the model evidence for
each of the six models. We can see that the best number of clusters is two, as
the minimum value of Laplace approximation to the negative log model
evidence for
DMM models as a function of k, determines the optimal k. The optimal k suggests
to fit a model with k mixtures of Dirichlet distributions as a prior.

```{r dmm2, fig.width=5, fig.height=3}
library(miaViz)
p <- plotDMNFit(altExp(tse, "prevalent"), type = "laplace", name = "DMM")
p + theme_minimal(base_size = 11)
```

The plot above suggests that two clusters describe the data the best. Now 
we can plot which taxa define the key cluster features.

```{r dmm3, fig.width=6, fig.height=2}
# Get the estimates on how much each phyla contributes on each cluster
best_model <- metadata(altExp(tse, "prevalent"))$DMM$dmm[2]
drivers <- as.data.frame(best_model[[1]]@fit$Estimate)

# Plot by utilizng miaViz::plotLoadings
plotLoadings(as.matrix(drivers), ncomponents = 2)
```

As well as in hierarchical clustering, we can also visualize the clusters by
projecting the data onto two-dimensional surface calculated using MDS. The plot
rather clearly demonstrates the cluster division.

```{r dmm4}
# DMM clusters
plotReducedDim(altExp(tse, "prevalent"), "MDS", theme_size = 13) +
              geom_point(size = 3, alpha = 0.6,
                      aes(color = altExp(tse, "prevalent")$dmmclust)) +
              labs(color = "DMMclust", title = "DMM") + theme_minimal()
```

## Dimensionality reduction

Dimensionality reduction can be considered as a part of community typing,
where the aim is to find underlying structure in the data. Dimensionality
reduction relies in the idea that the data can be describe in a
lower rank representation, where latent features with different loads of the
observed ones describe the sample set. This representation stores the
information about sample dissimilarity with significantly reduced amount of
data dimensions. Unlike clustering, where each sample is assigned into one
cluster, dimensionality reduction allows continuous memberships for all samples
into all community types. Dimensionality reduction techniques are explained in
chapter [@sec-community-similarity] in more detail.

### Non-negative matrix factorization

In this section, we will particularly focus on Non-negative Matrix Factorization
(NMF). NMF decomposes the original data *X* into two lower rank  matrices: *H*
representing key taxa characteristics of each
community type and *W* representing
continuous sample memberships across all community types. NMF algorithm
calculates the minimum distance between the original data and it's approximation
using Kullback-Leibler divergence.

Here, we use `mia` wrapper `getNMF`. To reduce calculation time,
we set the number of components `k` to four instead of optimizing the number of
components that describe the original data the best.
The matrix returned by `getNMF` contains the sample scores, while the NMF output
and NMF feature loadings are stored in the attributes `NMF_output` and 
`loadings`, respectively.

```{r nmf1}
library(NMF)

# Calculate NMF
set.seed(3221)
res <- getNMF(altExp(tse, "prevalent"), k = 4)

# Obtain the relative abundance of ES in each sample
wrel <- t(apply(res, 1, function (x) {x/sum(x)}))

# Define the community type of a sample with the highest relative abundance
altExp(tse, "prevalent")$NMFcomponent <- as.factor(apply(wrel, 1, which.max))
table(altExp(tse, "prevalent")$NMFcomponent)
```

Now, we can plot the NMF community types calculated based on maximum membership
for each sample onto a two-dimensional surface.

```{r nmf2}
# NMF community types
plotReducedDim(
    altExp(tse, "prevalent"), "MDS", colour_by = "NMFcomponent",
    point_size = 3, theme_size = 13) +

    labs(color = "NMFcomponent", title = "NMF") + theme_minimal()
```

## Heatmaps with clusters {#sec-clustered-heatmap}

In [@sec-composition-heatmap], we visualized community composition with
heatmaps. We can visualize the data similarly but add addition information on
clusters.

```{r pheatmap1}
library(ComplexHeatmap)

# Agglomerate to Genus level
tse <- agglomerateByPrevalence(tse, rank = "genus")

# Apply clr and scale data
tse <- transformAssay(tse, method = "clr", pseudocount = TRUE)
tse <-  transformAssay(
    tse, assay.type = "clr", method = "standardize", MARGIN = "rows")
# Gets the assay table
mat <- assay(tse, "standardize")
```

Now we can cluster rows.

```{r pheatmap2}
library(ape)
library(ggtree)

# Hierarchical clustering
tse <- addCluster(
    tse, assay.type = "standardize", HclustParam(method = "complete"),
    full = TRUE)
taxa_hclust <- metadata(tse)[["clusters"]][["hclust"]]
# Creates a phylogenetic tree
taxa_tree <- as.phylo(taxa_hclust)
ggtree(taxa_tree)
```

Based on the phylo tree, we decide to create eight clusters.

```{r pheatmap4}
# Hierarchical clustering with k = 3
tse <- addCluster(
    tse, assay.type = "standardize",
    HclustParam(method = "complete", cut.fun = cutree, k = 8))
```

For columns, we can utilize k-means clustering. First we have to specify the
number of clusters.



```{r pheatmap6}
library(factoextra)
library(NbClust)
library(patchwork)

# Elbow method
p1 <- fviz_nbclust(assay(tse, "standardize"), kmeans, method = "wss")
# Silhouette method
p2 <- fviz_nbclust(assay(tse, "standardize"), kmeans, method = "silhouette")

p1 + p2
```

Based on the results, we split the columns into six clusters.

```{r pheatmap7}
# Hierarchical clustering with k = 4
tse <- addCluster(
    tse, assay.type = "standardize", by = "cols",
    KmeansParam(centers = 6))

# Prints samples and their clusters
colData(tse)[["clusters"]]
```

Now we can create a heatmap with additional annotations.

```{r pheatmap8, fig.width=13, fig.height=7}
# Order the data based on clusters
tse <- tse[
    order(rowData(tse)[["clusters"]]),
    order(colData(tse)[["clusters"]])]
# Get data for plotting
mat <- assay(tse, "standardize")
taxa_clusters <- rowData(tse)[, "clusters", drop = FALSE] |>
  as.data.frame()
sample_data <- colData(tse)[, c("clusters", "Gender"), drop = FALSE] |>
  as.data.frame()
colnames(sample_data) <- c("clusters_col", "gender")

# Determines the scaling of colors
# Scale colors
breaks <- seq(
    -ceiling(max(abs(mat))), ceiling(max(abs(mat))),
    length.out = ifelse( max(abs(mat))>5, 2*ceiling(max(abs(mat))), 10 ) )
colors <- colorRampPalette(c("darkblue", "blue", "white", "red", "darkred"))(length(breaks)-1)

pheatmap(
    mat, annotation_row = taxa_clusters,
    annotation_col = sample_data,
    cluster_cols = FALSE, cluster_rows = FALSE,
    breaks = breaks,
    color = colors)
```

::: {.callout-tip}
## Additional Community Typing

For more community typing techniques applied to the 'SprockettTHData'
data set, see the attached .Rmd file.

Link:

   * [Rmd](extra_material/add-comm-typing.Rmd)

:::
