# Correlation {#sec-correlation}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

In correlation — or association analysis more generally — we can evaluate the
relationships between numeric variables. These variables can be taxa or
patient metadata. For instance, we might be interested in which taxa are
present simultaneously with others, or whether body weight is associated with
the abundance of certain taxa. In this chapter, we will demonstrate how to
perform correlation analysis with `getCrossAssociation()` method.

## Association between taxa

Here we demonstrate, how to analyse which bacteria co-exists in the dataset.

```{r}
#| label: association1

library(mia)

data("peerj13075")
tse <- peerj13075

# Agglomerate to certain taxonomy level
tse <- agglomerateByPrevalence(tse, rank = "class")

# Apply clr-transform and scale
tse <- transformAssay(tse, method = "clr", pseudocount = TRUE)
tse <- transformAssay(
    tse, assay.type = "clr", method = "standardize", MARGIN = "rows")

# Get correlation results
res <- getCrossAssociation(
    tse, tse, assay.type1 = "clr", assay.type2 = "clr",
    test.signif = TRUE, mode = "matrix")
```

We can visualize the result with heatmap as we do later in this chapter, or
we can visualize the results with correlation network plot as done below.

```{r}
#| label: association3

library(qgraph)

# Create correlation network plot
qgraph(
    res$cor, layout = "spring", labels = colnames(res$cor),
    label.cex = 1.2, theme = "colorblind",
    node.width = 1.5, node.height = 2)
```

You can find more on networks from [@sec-network-learning].

## Association between taxa and sample metadata

Now, we can calculate alpha diversity indices, and evaluate if they have
significant association with taxa.

```{r}
#| label: association4

# Calculate diversity measures
index <- c(
    "shannon", "log_modulo_skewness", "coverage", "inverse_simpson", "gini")
tse <- addAlpha(tse, index = index)

# Get correlation results
res <- getCrossAssociation(
    tse, tse, assay.type1 = "clr", col.var2 = index,
    test.signif = TRUE, mode = "matrix")
```

Below, we present the results using a heatmap visualization.

```{r}
#| label: association5
#| fig-width: 8
#| fig-height: 8

library(ComplexHeatmap)
library(shadowtext)

# Function for marking significant correlations with "X"
add_signif <- function(j, i, x, y, width, height, fill) {
    # If the p-value is under threshold
    if( !is.na(res$p_adj[i, j]) & res$p_adj[i, j] < 0.05 ){
        # Print "X"
        grid.shadowtext(
            sprintf("%s", "X"), x, y, gp = gpar(fontsize = 8, col = "#f5f5f5"))
    }
}

# Create a heatmap
p <- Heatmap(res$cor,
    # Print values to cells
    cell_fun = add_signif,
    heatmap_legend_param = list(
        title = "correlation", legend_height = unit(5, "cm")),
    column_names_rot = 45
    )
p
```

## Association between sample metadata variables {#sec-alpha-correlation}

Finally, we demonstrate how to calculate correlation between sample metadata
variables. Here we estimate correlation between alpha diversity measures.

Compared to the solution in [@sec-compare-alpha], `getCrossAssociation()`
allows us to calculate correlations in bulk easily, without the need for
looping.

```{r}
#| label: association6

# Get correlation results
res <- getCrossAssociation(
    tse, tse, col.var1 = index, col.var2 = index,
    test.signif = TRUE, mode = "matrix")

# Create a heatmap and store it
p <- Heatmap(res$cor,
    # Print values to cells
    cell_fun = add_signif,
    heatmap_legend_param = list(
        title = "correlation", legend_height = unit(5, "cm")),
    column_names_rot = 45
    )
p
```

::: {.callout-tip}
## Cross-association

See [@sec-cross-correlation] for further information on correlation and association analyses.
:::
