# Cross-association {#sec-cross-correlation}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

```{r load-pkg-data}
library(mia)
```

Multi-omics approaches integrate data from multiple sources. For
example, we can integrate taxonomic abundance profiles with
metabolomic or other biomolecular profiling data to observe
associations, make predictions, or aim at causal
inferences. Integrating evidence across multiple sources can lead to
enhanced predictions, more holistic understanding, or facilitate the
discovery of novel biomarkers. In the following chapters we demonstrate common
multi-assay data integration tasks.

As an example dataset, we use a data from the following publication:
[-@Hintikka2021] Xylo-Oligosaccharides in prevention of hepatic
steatosis and adipose tissue inflammation: associating taxonomic and
metabolomic patterns in fecal microbiota with biclustering.
In this study, mice were fed either with a high-fat or a low-fat diet,
and with or without prebiotics, for the purpose of studying whether prebiotics
attenuate the negative impact of a high-fat diet on health.

This example data can be loaded from the `mia` package. The data is
already in `MAE` format which is tailored for multi-assay analyses
(see [@sec-containers]). The dataset includes three different experiments:
microbial abundance data, metabolite concentrations, and data about
different biomarkers. If you like to construct the same data object from the
original files instead, [here]([@sec-loading-experimental-microbiome-data]) 
you can find help for importing data into an `SE`object.

## Cross-association Analysis {#sec-cross-correlation}

Cross-association analysis is a straightforward approach that can
reveal strength and type of associations between data sets. For instance,
we can analyze if higher presence of a specific taxon relates to higher
levels of a biomolecule. Correlation analyses within dataset were already
discussed in [@sec-correlation].

```{r cross-association1}
# Load the data
data(HintikkaXOData, package = "mia")
mae <- HintikkaXOData
```

```{r}
# Available alternative experiments
experiments(mae)
```

```{r}
# Microbiome data
getWithColData(mae, "microbiota")
```

```{r cross-association3}
# Metabolite data
getWithColData(mae, "metabolites")
```

```{r cross-association4}
# Biomarker data
getWithColData(mae, "biomarkers")
```

Next we can perform a cross-association analysis. Let us analyze if
individual bacteria genera are correlated with concentrations of
individual metabolites. This helps to answer the following question: "If
bacterium X is present, is the concentration of metabolite Y lower or higher"?

```{r cross-correlation5}
# Agglomerate microbiome data at family level
mae[[1]] <- agglomerateByPrevalence(mae[[1]], rank = "Family", na.rm = TRUE)
# Does log10 transform for microbiome data
mae[[1]] <- transformAssay(mae[[1]], method = "log10", pseudocount = TRUE)

# Give unique names, so that we do not have problems when we are creating a plot
rownames(mae[[1]]) <- getTaxonomyLabels(mae[[1]])

# Cross correlates data sets
res <- getCrossAssociation(
    mae,
    experiment1 = 1,
    experiment2 = 2,
    assay.type1 = "log10",
    assay.type2 = "nmr",
    method = "spearman",
    test.signif = TRUE,
    p_adj_threshold = NULL,
    cor_threshold = NULL,
    # Remove when mia is fixed
    mode = "matrix",
    sort = TRUE,
    show.warnings = FALSE)
```

Next, we create a heatmap depicting all cross-correlations between bacterial
genera and metabolite concentrations.

```{r cross-correlation6, fig.width=10, fig.height=8}
library(ComplexHeatmap)
library(shadowtext)

# Function for marking significant correlations with "X"
add_signif <- function(j, i, x, y, width, height, fill) {
    # If the p-value is under threshold
    if( !is.na(res$p_adj[i, j]) & res$p_adj[i, j] < 0.05 ){
        # Print "X"
        grid.shadowtext(
            sprintf("%s", "X"), x, y, gp = gpar(fontsize = 8, col = "#f5f5f5"))
    }
}

# Create a heatmap
p <- Heatmap(res$cor,
    # Print values to cells
    cell_fun = add_signif,
    heatmap_legend_param = list(
        title = "correlation", legend_height = unit(5, "cm")),
    column_names_rot = 45
    )
p
```
