# Import {#sec-import}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
library(DT)
chapterPreamble()
```

This chapter demonstrates how data can be imported into data containers from
files. Additionally, multiple databases offer curated microbiome data, which
are introduced in [@sec-example-data].

In prior to importing the data to container, the raw sequences must be
mapped to abundances. The specific method depends on sequencing type and other
data-specific factors. There are several resources available to guide this
process [Metagenomics wiki](https://www.metagenomics.wiki/) being one of them.
In [@sec-16s-workflow] we provide an example workflow for processing 16S rRNA
sequences into a data container using DADA2.

## Import microbiome data from files {#sec-loading-experimental-microbiome-data}

The data containers can be constructed from scratch. For most common microbiome
data formats, there is dedicated importers available which streamlines the
importing.

### Standard data formats

Specific import functions are provided for:

- BIOM files (see `help(mia::importBIOM)`)
- QIIME2 files (see `help(mia::importQIIME2)`)
- Mothur files (see `help(mia::importMothur)`)
- MetaPhlAn files (see `help(mia::importMetaPhlAn)`)
- HUMAnN files (see `help(mia::importHUMAnN)`)
- taxpasta files (see `help(mia::importTaxpasta)`)

Here we show how [Biom files](https://biom-format.org/) are imported into
a `TreeSE` object using as an example Tengeler2020, which is further described
[here](https://microbiome.github.io/mia/reference/Tengeler2020.html).
This dataset consists of 3 files, which can be
fetched or downloaded from
[this repository](https://github.com/microbiome/data/tree/main/Tengeler2020):

-   biom file: abundance table and taxonomy information
-   csv file: sample metadata
-   tree file: phylogenetic tree

To begin with, we store the data in a local directory within the working
directory, such as _data/_, and define the source file paths.

```{r}
biom_file_path <- system.file(
    "extdata", "Aggregated_humanization2.biom", package = "OMA")
sample_meta_file_path <- system.file(
    "extdata", "Mapping_file_ADHD_aggregated.csv", package = "OMA")
tree_file_path <- system.file(
    "extdata", "Data_humanization_phylo_aggregation.tre", package = "OMA")
```

Now we can read in the biom file and convert it into a `TreeSE` object. In
addition, we retrieve the rank names from the prefixes of the feature names
and then remove them with the `rank.from.prefix` and `prefix.rm` optional
arguments.

```{r}
library(mia)

# read biom and convert it to TreeSE
tse <- importBIOM(
    biom_file_path,
    rank.from.prefix = TRUE,
    prefix.rm = TRUE)

# Check
tse
```

The `assays` slot includes a list of abundance tables. The imported
abundance table is named as "counts".  Let us inspect only the first
cols and rows.

```{r}
assay(tse, "counts")[1:3, 1:3]
```

The `rowdata` includes taxonomic information from the biom file. The `head()`
command shows just the beginning of the data table for an overview.

`knitr::kable()` helps print the information more nicely.

```{r}
head(rowData(tse))
```

We further polish the feature names by removing unnecessary characters and
then replace the original `rowData` with its updated version.

```{r}
# Genus level has additional '\"', so let's delete that also
rowdata_modified <- BiocParallel::bplapply(
    rowData(tse),
    FUN = stringr::str_remove,
    pattern = '\"')

# rowdata_modified is a list, so convert this back to DataFrame format.
# and assign the cleaned data back to the TSE rowData
rowData(tse) <- DataFrame(rowdata_modified)

# Now we have a nicer table
head(rowData(tse))
```

We notice that the imported biom file did not contain any `colData` yet,
so only an empty dataframe appears in this slot.

```{r}
head(colData(tse))
```

Let us add `colData` from the sample metadata, which is stored in a CSV file.

```{r}
# CSV file with colnames in the first row and rownames in the first column
sample_meta <- read.csv(
    sample_meta_file_path, sep = ",", row.names = 1)

# Add this sample data to colData of the taxonomic data object
# Note that the data must be given in a DataFrame format
colData(tse) <- DataFrame(sample_meta)
```

Now the `colData` includes the sample metadata.

```{r}
head(colData(tse))
```

Finally, we add a phylogenetic tree to the `rowData` slot. Such feature is
available only in `TreeSE` objects. Similarly, Trees specifying the sample
hierarchy can be stored in the `colTree` slot.

Here, we read in the file containing the phylogenetic tree and insert it
in corresponding slot of the `TreeSE` object.

```{r}
# Reads the tree file
tree <- ape::read.tree(tree_file_path)

# Add tree to rowTree
rowTree(tse) <- tree

# Check
tse
```

Now the `rowTree` slot contains the phylogenetic tree:

```{r, eval=FALSE}
head(rowTree(tse))
```

### Non-standard formats {#sec-import-from-file}

Microbiome (taxonomic) profiling data is commonly distributed in
various file formats. You can import such external data files as a
`TreeSE` object, but the details depend on the file
format. Here, we provide examples for common formats. Some datasets and raw
files to learn how to import raw data and construct `TreeSE/MAE` containers are
available in the
[microbiome data repository](https://github.com/microbiome/data).

#### CSV import

**CSV data tables** can be imported with the standard R functions,
then converted to the desired format. For detailed examples, you can
check the
[Bioconductor course material](https://bioconductor.org/help/course-materials/2019/BSS2019/04_Practical_CoreApproachesInBioconductor.html)
by Martin Morgan. You can also check the [example
files](https://github.com/microbiome/OMA/tree/master/data) and
construct your own CSV files accordingly.

Recommendations for the CSV files are the following. File names are
arbitrary; we refer here to the same names as in the examples:

- Abundance table (`assay_taxa.csv`): data matrix (features x
samples); first column provides feature IDs, the first row provides
sample IDs; other values should be numeric (abundances).

- Row data (`rowdata_taxa.csv`): data table (features x info); first
column provides feature IDs, the first row provides column headers;
this file usually contains the taxonomic mapping between different
taxonomic levels. Ideally, the feature IDs (row names) match one-to-one with
the abundance table row names.

- Column data (`coldata.csv`): data table (samples x info); first
column provides sample IDs, the first row provides column headers;
this file usually contains the sample metadata/phenodata (such as
subject age, health etc). Ideally, the sample IDs match one-to-one with
the abundance table column names.

After you have set up the CSV files, you can read them in R:

```{r importingcsv1, message=FALSE}
count_file  <- system.file("extdata", "assay_taxa.csv", package = "OMA")
tax_file    <- system.file("extdata", "rowdata_taxa.csv", package = "OMA")
sample_file <- system.file("extdata", "coldata.csv", package = "OMA")

# Load files
counts  <- read.csv(count_file, row.names=1)   # Abundance table (e.g. ASV data; to assay data)
tax     <- read.csv(tax_file, row.names=1)     # Taxonomy table (to rowData)
samples <- read.csv(sample_file, row.names=1)  # Sample data (to colData)
```

After reading the data in R, ensure the following:

- abundance table (`counts`): numeric `matrix`, with feature IDs as
  rownames and sample IDs as column names.

- `rowdata` (`tax`): `DataFrame`, with feature IDs as rownames. If this
  is a `data.frame` you can use the function `DataFrame()` to change
  the format. Column names are free but in microbiome analysis they
  usually they refer to taxonomic ranks. The rownames in rowdata
  should match with rownames in abundance table.

- `coldata` (`samples`): `DataFrame`, with sample IDs as rownames. If
  this is a `data.frame` you can use the function `DataFrame()` to
  change the format.  Column names are free. The rownames in `coldata`
  should match with colnames in abundance table.

**Always ensure that the tables have rownames!** The _TreeSE_ constructor
compares  rownames and ensures that, for example, right samples are linked
with right patient.

Also, ensure that the row and column names match one-to-one between
abundance table, `rowdata`, and `coldata`:

```{r importingcsv2}
# Match rows and columns
counts <- counts[rownames(tax), rownames(samples)]

# Let's ensure that the data is in correct (numeric matrix) format:
counts <- as.matrix(counts)
```

If you hesitate about the format of the data, you can compare to one
of the available demonstration datasets, and make sure that your data
components have the same format.

There are many different source files and many different ways to read
data in R. One can do data manipulation in R as well. Investigate the
entries as follows.

```{r demodata, message=FALSE}
# coldata rownames match assay colnames
all(rownames(samples) == colnames(counts)) # our dataset
class(samples) # should be data.frame or DataFrame

# rowdata rownames match assay rownames
all(rownames(tax) == rownames(counts)) # our dataset
class(tax) # should be data.frame or DataFrame

# Counts
class(counts) # should be a numeric matrix
```

::: {.callout-important}
## Important!

Ensure that colnames of assay match with rownames of `colData`, and rownames of
assay match with rownames of `rowData`.

If your data do not have names, you have to be especially careful, since this
can lead to errors!
:::

#### Constructing TreeSE

Now, let's create the `TreeSE` object from the input data tables. Here
we also convert the data objects in their preferred formats:

   - counts --> numeric matrix
   - rowData --> DataFrame
   - colData --> DataFrame

The `SimpleList` could be used to include multiple alternative assays, if
necessary.

```{r importingcsv3}
# Create a TreeSE
tse_taxa <- TreeSummarizedExperiment(
    assays =  SimpleList(counts = counts),
    colData = DataFrame(samples),
    rowData = DataFrame(tax))

tse_taxa
```

Now you should have a ready-made `TreeSE` data object that can be used in
downstream analyses.

#### Constructing MAE

To construct a `MAE` object, just combine multiple `TreeSE` data containers. 

Here we import metabolite data from the same study.

```{r importingcsv4, message=FALSE}
count_file <- system.file("extdata", "assay_metabolites.csv", package = "OMA")
sample_file <- system.file("extdata", "coldata.csv", package = "OMA")

# Load files
counts  <- read.csv(count_file, row.names=1)
samples <- read.csv(sample_file, row.names=1)

# Create a TreeSE for the metabolite data
tse_metabolite <- TreeSummarizedExperiment(
    assays = SimpleList(concs = as.matrix(counts)),
    colData = DataFrame(samples))

tse_metabolite
```

::: {.callout-important}
## Important!

When creating TreeSE, `assay` must be a matrix, and both `colData` and `rowData` must
be DataFrame objects.
:::

Now we can combine these two experiments into _MAE_.

```{r importingcsv5}
# Create an ExperimentList that includes experiments
experiments <- ExperimentList(
    microbiome = tse_taxa, metabolite = tse_metabolite)

# Create a MAE
mae <- MultiAssayExperiment(experiments = experiments)

mae
```

## Data resources {#sec-example-data}

Open demonstration data for testing and benchmarking purposes is
available from multiple locations. This chapter introduces some
options. The other chapters of this book provide ample examples about
the use of the data.

### Package data {#sec-package-data}

The `mia` R package contains example datasets that are direct
conversions from the alternative `phyloseq` container to the
`TreeSE` container.

List the [available
datasets](https://microbiome.github.io/mia/reference/mia-datasets.html) in
the `mia` package:

```{r, message=FALSE}
library(mia)
data(package="mia")
```

Load the `GlobalPatterns` data from the `mia` package:

```{r, message=FALSE}
data("GlobalPatterns", package="mia")
GlobalPatterns
```

R packages contain additional demonstration data sets (see the Datasets
section of the reference page):

* [mia reference](https://microbiome.github.io/mia/reference/index.html)
* [miaViz reference](https://microbiome.github.io/miaViz/reference/index.html)
* [miaTime reference](https://microbiome.github.io/miaTime/reference/index.html)

### ExperimentHub data

[ExperimentHub](https://bioconductor.org/packages/release/bioc/vignettes/ExperimentHub/inst/doc/ExperimentHub.html)
provides a variety of data resources, including the
[microbiomeDataSets](https://bioconductor.org/packages/release/data/experiment/html/microbiomeDataSets.html)
package [@Morgan2021; @microlahti2021].

A table of the available datasets is available through the
`availableDataSets()` function.

```{r, message=FALSE}
library(microbiomeDataSets)
availableDataSets()
```

All data are downloaded from ExperimentHub and cached for local
re-use. Check the [man pages of each
function](https://microbiome.github.io/microbiomeDataSets/reference/index.html)
for a detailed documentation of the data contents and references. Let
us retrieve a `r Biocpkg("MAE")` dataset:

```{r eval=FALSE, message=FALSE}
# mae <- HintikkaXOData()
# Since HintikkaXOData is now added to mia, we can load it directly from there
# We suggest to check other datasets from microbiomeDataSets
data(HintikkaXOData, package = "mia")
mae <- HintikkaXOData
```

Data is available in `r Biocpkg("SE")`, `r Biocpkg("TreeSE")` and
`r Biocpkg("MAE")` data containers; see the for example [@sec-cross-correlation]
for more details.

### Curated metagenomic data

[curatedMetagenomicData](https://bioconductor.org/packages/release/data/experiment/html/curatedMetagenomicData.html)
is a large collection of curated human microbiome datasets, provided as
`TreeSE` objects [@Pasolli2017]. The resource
provides curated human microbiome data including gene families, marker
abundance, marker presence, pathway abundance, pathway coverage, and
relative abundance for samples from different body sites. See the
package homepage for more details on data availability and access.

As one example, let us retrieve the Vatanen (2016) [@Vatanen2016] data
set. This is a larger collection with a bit longer download time.

```{r, message=FALSE, eval=FALSE}
library(curatedMetagenomicData)
tse <- curatedMetagenomicData("Vatanen*", dryrun = FALSE, counts = TRUE)
```

###  Human microbiome compendium

[MicroBioMap](https://seandavi.github.io/MicroBioMap/) dataset includes
over 170k samples of publicly available 16S rRNA amplicon sequencing data,
all processed using the same pipeline and reference database [@microbiomap].
After installing the MicroBioMap package (see the
[original website](https://github.com/seandavi/MicroBioMap#microbiome-compendium)
for instructions), you can load the compendium with

```{r, message=FALSE, eval=FALSE}
library(MicroBioMap)
cpd <- getCompendium()
```

This returns a `TreeSE` object. Currently,
the `rowTree` slot of the `TreeSE` is not populated.

After loading the compendium, you will have immediate access to
nearly 170,000 microbiome samples of publicly available 16S rRNA amplicon
sequencing data, all processed using the same pipeline and reference database.
For more use examples in R/Bioconductor, see the
[MicroBioMap vignette](https://seandavi.github.io/MicroBioMap/articles/overview.html).

### Other data sources

The current collections provide access to vast microbiome data
resources. The output has to be converted into `TreeSE/MAE` separately.

- [MGnifyR](https://bioconductor.org/packages/release/bioc/html/MGnifyR.html) provides access to [EBI/MGnify](https://www.ebi.ac.uk/metagenomics/)
- [HoloFoodR](https://bioconductor.org/packages/release/bioc/html/HoloFoodR.html) provides access to [EBI/HoloFood](https://www.holofooddata.org/)
- [qiitr](https://github.com/cran/qiitr) provides access to [QIITA](https://qiita.com/about)
- [qiime2R](https://github.com/microbiome/qiime2R) provides access to [QIIME2](https://docs.qiime2.org/2024.2/)
