# Ordination-based multiassay analysis {#sec-multiassay_ordination}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

In this chapter, we continue from where we left off in the previous section
(referenced as [@sec-cross-correlation]). Specifically, we introduce an
analytical method called Multi-Omics Factor Analysis. This method is described
as "ordination-based," meaning it involves techniques that reduce the
dimensionality of the data while preserving as much variability as possible.
Ordination methods are often used in multivariate analysis to visualize
patterns, trends, and relationships in complex data sets, making them
particularly useful for multi-omics data where multiple types of biological
data are integrated and analyzed together.

See [@sec-community-similarity] for more information on ordination methods.

## Multi-Omics Factor Analysis {#sec-mofa}

Multi-Omics Factor Analysis (MOFA) is an unsupervised method for integrating
multi-omic data sets in a downstream analysis [@Argelaguet2018]. It could be
seen as a generalization of principal component analysis. Yet, with the ability
to infer a latent (low-dimensional) representation, shared among the multiple
(-omics) data sets in hand.

We use the R [MOFA2](https://biofam.github.io/MOFA2/index.html)
package for the analysis, and
[install](https://biofam.github.io/MOFA2/installation.html) the
corresponding dependencies.

```{r mofa_data1}
library(mia)
# Load the data
data(HintikkaXOData, package = "mia")
mae <- HintikkaXOData
```

The `mae` object could be used straight to create the MOFA model. Yet,
we transform our assays since the model assumes normality per
default, and Gaussian model is recommended
(see [MOFA2 FAQ](https://biofam.github.io/MOFA2/faq.html)). However, Poisson and
Bernoulli distribution models are also offered.

Note that duplicates, such as "uncultured", might appear when aggregating the
microbiome data by a taxonomic rank. To check for duplicates, run
`any(duplicated(rownames(mae[[1]])))`. If it returns `TRUE`, then the
duplicates are present. We can add
`rownames(mae[[1]]) <- getTaxonomyLabels(mae[[1]], make.unique=TRUE)` to
remove them.

```{r, message=FALSE, warning=FALSE}
#| label: mofa3

library(MOFA2)
# For simplicity, classify all high-fat diets as high-fat, and all the low-fat
# diets as low-fat diets
colData(mae)$Diet <- ifelse(
    colData(mae)$Diet == "High-fat" | colData(mae)$Diet == "High-fat + XOS",
    "High-fat", "Low-fat")

# Agglomerate microbiome data
mae[[1]] <- agglomerateByPrevalence(mae[[1]], rank = "Genus")
# Transforming microbiome data with clr and by scaling
mae[[1]] <- transformAssay(mae[[1]], method = "clr", pseudocount = TRUE)
mae[[1]] <- transformAssay(
    mae[[1]], assay.type = "clr", method = "standardize", MARGIN = "rows")

# Transforming metabolomic data with log10 and by scaling
mae[[2]] <- transformAssay(mae[[2]], assay.type = "nmr", method = "log10")
mae[[2]] <- transformAssay(
    mae[[2]], assay.type = "log10", method = "standardize")

# Transforming biomarker data by scaling
mae[[3]] <- transformAssay(
    mae[[3]], assay.type = "signals", method = "standardize", MARGIN = "rows")

# Removing the assays no longer needed
assays(mae[[1]]) <- assays(mae[[1]])["standardize"]
assays(mae[[2]]) <- assays(mae[[2]])["standardize"]
assays(mae[[3]]) <- assays(mae[[3]])["standardize"]

# Building our mofa model
model <- create_mofa_from_MultiAssayExperiment(
    mae,
    groups = "Diet",
    extract_metadata = TRUE)
model
```

Model options can be defined as follows:

```{r, message=FALSE, warning=FALSE}
#| label: mofa4

model_opts <- get_default_model_options(model)
model_opts$num_factors <- 5
model_opts |> head()
```

Training options for the model are defined in the following way:

```{r, message=FALSE, warning=FALSE}
#| label: mofa5

train_opts <- get_default_training_options(model)
train_opts |> head()
```

The model is then prepared  with `prepare_mofa()` and trained with `run_mofa()`:

```{r, message=FALSE, warning=FALSE}
#| label: mofa6

model <- prepare_mofa(
  object = model,
  model_options = model_opts
)

# Some systems may require the specification `use_basilisk = TRUE`
# so it has been added to the following code
model <- run_mofa(model, use_basilisk = TRUE)
```

The explained variance is visualized with the `plot_variance_explained()` function.

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=10}
#| label: mofa7

library(patchwork)
library(ggplot2)

plot_list <- plot_variance_explained(
    model,
    x = "view", y = "factor",
    plot_total = TRUE)

wrap_plots(plot_list, nrow = 2) +
    plot_annotation(
        title = "Variance Explained per factor and assay",
        theme = theme(plot.title = element_text(hjust = 0.5)))
```

From the plot, we can observe that the microbiota accounts for most of the
variability in the data. Biomarkers do not significantly explain the
variability. The variability in the microbiota is primarily captured in factors
1 and 2, while metabolites are primarily represented in factor 2.

We can then visualize the top weights for microbiota and metabolites for
the first two factors to analyze co-varying features.

```{r, warning=FALSE, message=FALSE, fig.height=10, fig.width=10}
#| label: mofa8

plot_list <- lapply(
    c("microbiota", "metabolites"),
    plot_top_weights,
    object = model,
    factors = c(1, 2),
    nfeatures = 10
    )

wrap_plots(plot_list, ncol = 1) & theme(text = element_text(size = 8))
```

From the visualization, we can see that glycerol, lactate, and choline are
positively associated with factor 2, while acetate is negatively associated
among the metabolites. In terms of microbiota, Ruminococcaceae and
Ruminodostridium, for instance, are positively associated with the same factor.
This indicates that these features co-vary in the data, suggesting a positive
association with glycerol, lactate, and choline, and a negative association
with acetate and these microbes.

More tutorials and examples of using the package are found at
[MOFA2 tutorials](https://biofam.github.io/MOFA2/tutorials.html).
