# Subsetting {#sec-treese_subsetting}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

In this chapter, we explore the concept of subsetting. Subsetting or
filtering is the process of selecting specific rows or columns from a
dataset based on certain criteria. When you subset your data, you retain the
original values of the selected data points. For example, if you have a
dataset of microbial taxa and you choose to keep only certain species, you
still have the individual counts for those species. This allows you to analyze
the data in detail, but you may lose information about other species that are
not included.

Subsetting data helps to draw the focus of an analysis on particular
sets of samples and / or features. When dealing with large datasets,
the subset of interest can be extracted and investigated
separately. This might improve performance and reduce the
computational load.

Let us store `GlobalPatterns` into `tse` and check its original number of
features (rows) and samples (columns).

```{r}
#| label: load_data

# Load libraries and data
library(mia)
library(dplyr)
library(knitr)

data("GlobalPatterns", package = "mia")
tse <- GlobalPatterns

dim(tse)
```

::: {.callout-note}
## Note

When subsetting by sample, expect the number of columns to decrease. When 
subsetting by feature, expect the number of rows to decrease.
:::

## Subset by sample (column-wise)

Several criteria can be used to subset by sample:

- origin
- sampling time
- sequencing method
- DNA / RNA barcode
- cohort

For the sake of demonstration, here we will extract a subset containing only the
samples of human origin (feces, skin or tongue), stored as `SampleType` within
`colData(tse)` as well as in `tse`.

First, we would like to see all the possible values that `SampleType` can have
and how frequent those are:

```{r}
#| label: show_data

# Show the frequency of each value
tse$SampleType |> table() |> kable()
```

Next, we _logical index_ across the columns of `tse` (make sure to
leave the first index empty to select all rows) and filter for the
samples of human origin. For this, we use the information on the
samples from the meta data `colData(tse)`.

```{r}
#| label: filter_samples

# Subset by sample
tse_sub <- tse[ , tse$SampleType %in% c("Feces", "Skin", "Tongue")]

# Show dimensions
dim(tse_sub)
```

::: {.callout-note}
## Note

After subsetting, expect the number of columns to equal the
sum of the frequencies of the samples that you are interested
in. For instance, `ncols = Feces + Skin + Tongue = 4 + 3 + 2 = 9`.
:::

## Subset by feature (row-wise)

Similarly, here we will extract a subset containing only the features
that belong to the phyla Actinobacteria and Chlamydiae, stored as
`Phylum` within `rowData(tse)`. However, subsetting by feature implies
a few more obstacles, such as the presence of `NA` elements and the
possible need for agglomeration.

As previously, we would first like to see all the possible values that `Phylum`
can have and how frequent those are:

```{r}
#| label: show_phyla

# Show the frequency of each value
rowData(tse)$Phylum |> table() |> head() |> kable()
```

::: {.callout-note}
## Note

After subsetting, expect the number of columns to equal the
sum of the frequencies of the feature(s) that you are interested
in. For instance, `nrows = Actinobacteria + Chlamydiae = 1631 + 21 = 1652`.
:::

Depending on your research question, you might or might not need to
agglomerate the data in the first place: if you want to find the
abundance of each and every feature that belongs to Actinobacteria and
Chlamydiae, agglomeration is not needed. However, if you want to find the total
abundance of all features that belong to Actinobacteria or
Chlamydiae, agglomeration is recommended (see
[@sec-agglomeration] for details).

Next, we _logically index_ across the rows of `tse` (make sure to leave
the second index empty to select all columns) and filter for the
features that fall in either Actinobacteria or Chlamydiae group. For this,
we use the information on the samples from the metadata
`rowData(tse)`.

The first term with the `%in%` operator includes all the features
of interest, whereas the second term after the AND operator `&`
filters out all features that have an `NA` in place of the phylum variable.

```{r}
#| label: subset_features

# Subset by feature
selected <- rowData(tse)$Phylum %in% c("Actinobacteria", "Chlamydiae") &
  !is.na(rowData(tse)$Phylum)
tse_sub <- tse[selected, ]

# Show dimensions
dim(tse_sub)
```

## Subset by samples and features

Finally, we can subset data by sample and feature at once. The
resulting subset contains all the samples of human origin and all the
features of phyla Actinobacteria or Chlamydiae.

```{r}
#| label: subset_samples_and_feat

# Subset by sample and feature and remove NAs
selected_rows <- rowData(tse)$Phylum %in% c("Actinobacteria", "Chlamydiae") &
      !is.na(rowData(tse)$Phylum)
selected_cols <- tse$SampleType %in% c("Feces", "Skin", "Tongue")
tse_sub <- tse[selected_rows, selected_cols]

# Show dimensions
dim(tse_sub)
```

::: {.callout-note}
## Note

The dimensions of `tse_sub` are consistent with the previous subsets
(9 columns filtered by sample and 1652
rows filtered by feature).
:::

If a study was to consider and quantify the presence of Actinobacteria
as well as Chlamydiae in different sites of the human body,
`tse_sub` might be a suitable subset to start with.

## Filtering out empty samples

Sometimes data might contain, e.g., features that are not present in any of
the samples. This can occur, for example, after data subsetting. In certain
analyses, we might want to remove those instances. In this example, we are
interested only those features that belong to Species _Achromatiumoxaliferum_.

```{r}
#| label: select_archaea

ind <- rowData(tse)$Species == "Achromatiumoxaliferum"
ind[is.na(ind)] <- FALSE
tse_sub <- tse[ind, ]
```

Then we can calculate, how many times each feature was detected in each
samples.

```{r}
#| label: subset_empty

library(scuttle)

tse_sub <- addPerCellQCMetrics(tse_sub)
# List total counts of each sample
tse_sub$total |> head()
```

Now we can see that certain samples do not include any bacteria. We can remove
those.

```{r}
#| label: subset_empty2

# Remove samples that do not contain any bacteria
tse_sub <- tse_sub[ , tse_sub$total != 0]
tse_sub
```

We have now subsetted the dataset so that it only includes samples containing
the selected features. A similar approach can also be applied to filter features
based on the samples they appear in, ensuring both dimensions of the data are
relevant to your analysis.

## Filtering out zero-variance features

In some datasets, certain features remain constant across all samples—they show
no variance. If our goal is to study or model microbial differences between
groups, these zero-variance features hold no value. For a feature to reflect
differences between groups, it must exhibit variability.

By removing these invariant features, we can sharpen our focus on the more
informative features. This not only reduces the number of comparisons but also
helps our machine learning models learn from meaningful data without additional
noise.

To filter these features, we begin by calculating their standard deviation.
Then we visualize the variances with histogram.

```{r}
#| label: standard_deviation

# Calculate
rowData(tse)[["sd"]] <- rowSds(assay(tse, "counts"))
# Plot
hist(log(rowData(tse)[["sd"]]))
```

From the histogram of feature variances, we can establish a sensible threshold
for filtering. For example, using a threshold of 0 would effectively remove a
large set of features that show no variance.

It's important to note that the data is on a log scale, meaning that a
log-transformed value of 1 corresponds to 0 (i.e., log(1) = 0). This ensures
that features with zero variance are correctly identified and filtered out.

```{r}
#| label: filter_sd

th <- 1

selected <- rowData(tse)[["sd"]] > 1
tse_sub <- tse[selected, ]
tse_sub
```

Now the dataset includes the `r nrow(tse_sub)` features comapred to previous
`r nrow(tse)`. The dataset now includes the features that vary. Note that
this threshold choice depends on your research question, and this was a rather
conservative choice retaining most of the features.

After filtering, the dataset now contains `r nrow(tse_sub)` features, compared
to the original `r nrow(tse)` features. This means we are left with features
that exhibit variance. Keep in mind that the choice of threshold depends on the
specific research question, and in this case, we opted for a rather conservative
threshold that retains most features.

## Subset based on prevalence {#sec-subset_prev}

We can subset the data based on prevalence using `subsetByPrevalent()`, which
filters taxa that exceed a specified prevalence threshold, helping to remove
rare features that may be artefacts. Conversely, `subsetByRare()` allows us to
retain only taxa below the threshold, enabling a focus on rare features within
the dataset.

Here, we apply a prevalence threshold of 10% and a detection threshold of 1.
A feature is considered detected if it has at least one count in a sample, and
prevalent if it is detected in at least 10% of the samples. The function
`subsetByPrevalent()` retains the prevalent features, while `subsetByRare()`
keeps the features that are not prevalent.

```{r}
#| label: subset_by_rare

tse_sub <- subsetByRare(tse, rank = "Genus", prevalence = 0.1, detection = 1)
tse_sub
```

In some cases, agglomerating based on prevalence can provide more meaningful
insights. This process is illustrated in [@sec-agglomerate_prev].
