# Taxonomic Information {#sec-taxonomic-information}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

```{r, message=FALSE}
library(mia)
data("GlobalPatterns", package = "mia")
tse <- GlobalPatterns
```

Taxonomic information is a key part of analyzing microbiome data, and without
it, any type of data analysis probably will not make much sense. However,
the degree of detail of taxonomic information differs depending on the dataset
and annotation data used.

Therefore, the `mia` package expects a loose assembly of taxonomic information
and assumes certain key aspects:

* Taxonomic information is given as character vectors or factors in the
`rowData` of a `SummarizedExperiment` object.
* The columns containing the taxonomic information must be named `domain`,
`kingdom`, `phylum`, `class`, `order`, `family`, `genus`, `species` or with
a capital first letter.
* The columns must be given in the order shown above.
* column can be omitted, but the order must remain

In this chapter, we will refer to the co-abundant groups as CAGs, which are
clusters of taxa that co-vary across samples.

## Assigning taxonomic information

There are a number of methods to assign taxonomic information. We like to give
a short introduction about the methods available without ranking one over the
other. This has to be your choice based on the result for the individual
dataset.

### DADA2

The `dada2` package [@Callahan2016dada2] implements the `assignTaxonomy()`
function, which takes as input the ASV sequences associated with each
row of data and a training dataset. For more information visit the
[dada2 homepage](https://benjjneb.github.io/dada2/assign.html).

### DECIPHER

The `DECIPHER` package [@R_DECIPHER] implements the `IDTAXA` algorithm to assign
either taxonomic information or function information. For `mia`,
only the first option is of interest for now and more information can be
found on the [DECIPHER website](http://www2.decipher.codes/Classification.html).

## Functions to access taxonomic information

### Check taxonomy ranks in data

`checkTaxonomy()` checks whether the taxonomic information is usable for `mia`

```{r}
checkTaxonomy(tse)
```

Since the `rowData` can contain other data, `taxonomyRanks()` will return the 
columns that `mia` assumes to contain the taxonomic information.

```{r}
taxonomyRanks(tse)
```

This can then be used to subset the `rowData` to columns needed.

```{r}
rowData(tse)[, taxonomyRanks(tse)]
```

`taxonomyRankEmpty()` checks for empty values in the given `rank` and returns a 
logical vector of `length(x)`.

```{r}
all(!taxonomyRankEmpty(tse, rank = "Kingdom"))
table(taxonomyRankEmpty(tse, rank = "Genus"))
table(taxonomyRankEmpty(tse, rank = "Species"))
```

### Get taxonomy labels

`getTaxonomyLabels()` is a multi-purpose function, which turns taxonomic
information into a character vector of `length(x)`

```{r}
getTaxonomyLabels(tse) |> head()
```

By default, this will use the lowest non-empty information to construct a
string with the following scheme `level:value`. If all levels are the same,
this part is omitted, but can be added by setting `with.rank = TRUE`.

```{r}
phylum <- !is.na(rowData(tse)$Phylum) &
    vapply(data.frame(apply(
       rowData(tse)[, taxonomyRanks(tse)[3:7]], 1L, is.na)), all, logical(1))
getTaxonomyLabels(tse[phylum,]) |> head()
getTaxonomyLabels(tse[phylum,], with.rank = TRUE) |> head()
```

By default the return value of `getTaxonomyLabels()` contains only
unique elements by passing it through `make.unique`. This step can be
omitted by setting `make.unique = FALSE`.

```{r}
getTaxonomyLabels(tse[phylum,], with.rank = TRUE, make.unique = FALSE) |> head()
```

To apply the loop resolving function `resolveLoop()` from the
`TreeSummarizedExperiment` package [@R_TreeSummarizedExperiment] within
`getTaxonomyLabels()`, set `resolve.loops = TRUE`.

### Get information on certain taxa

The function `getUnique()` gives a list of unique taxa for the
specified taxonomic rank.

```{r}
getUnique(tse, rank = "Phylum") |> head()
```

With `mapTaxonomy()`, you can search information on certain taxa from the taxonomy
table. For instance, we can check all the taxa that matches with "Escherichia".

```{r}
#| label: mapTaxonomy
mapTaxonomy(GlobalPatterns, taxa = "Escherichia")
```


### Generate a hierarchy tree on the fly {#sec-fly-tree}

A hierarchy tree shows mapping between the taxonomic levels in taxonomic rank
table (included in `rowData`), rather than the detailed phylogenetic relations.
Usually, a phylogenetic tree refers to latter which is why we call here the
generated tree as "hierarchy tree".

To create a hierarchy tree, `getHierarchyTree()` used the information and
returns a `phylo` object. Duplicate information from the `rowData` is removed.

```{r}
getHierarchyTree(tse)
```

```{r}
tse <- addHierarchyTree(tse)
tse
```

The implementation is based on the `toTree()` function from the
`TreeSummarizedExperiment` package [@R_TreeSummarizedExperiment].

## Set taxonomy ranks

If your data includes taxonomy ranks that are not included by default in `mia`,
you can set the ranks manually. By doing so, `mia` will be able to detect and
utilize these taxonomy ranks from your data as expected.

Get default ranks of `mia`.

```{r}
#| label: get_default_ranks

getTaxonomyRanks()
```

Set ranks to your own ranks. Remember that the order is meaningful.

```{r}
#| label: get_ranks
#| echo: false
#| message: false

ranks <- getTaxonomyRanks()
```

```{r}
#| label: set_ranks

# Set ranks
setTaxonomyRanks(c("test", "test2", "apple"))

# Get ranks
getTaxonomyRanks()
```

```{r}
#| label: set_default_ranks
#| echo: false
#| message: false

# Set default ranks back
setTaxonomyRanks(ranks)
```
