# Data Wrangling {#sec-datawrangling}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

This chapter covers various basic techniques for preparing data for subsequent
analysis. These include subsetting, splitting, and modifying data, as well as
converting data to a `data.frame` through melting and merging multiple
`SummarizedExperiment` objects. See [@sec-containers] for basics of `TreeSE`.

## Subsetting {#sec-treese_subsetting}

**Subsetting** data helps to draw the focus of an analysis on particular
sets of samples and / or features. When dealing with large datasets,
the subset of interest can be extracted and investigated
separately. This might improve performance and reduce the
computational load.

Load:

* mia
* dplyr
* knitr
* data `GlobalPatterns`

```{r include = FALSE}
# Load libraries and data
library(mia)
library(dplyr)
library(knitr)
```

Let us store `GlobalPatterns` into `tse` and check its original number of
features (rows) and samples (columns). **Note**: When subsetting by sample,
expect the number of columns to decrease. When subsetting by feature, expect
the number of rows to decrease.

```{r}
# Store data into se and check dimensions
data("GlobalPatterns", package="mia")
tse <- GlobalPatterns
# Show dimensions (features x samples)
dim(tse)
```

### Subset by sample (column-wise)

For the sake of demonstration, here we will extract a subset containing only the
samples of human origin (feces, skin or tongue), stored as `SampleType` within
`colData(tse)` as well as in `tse`.

First, we would like to see all the possible values that `SampleType` can have
and how frequent those are:

```{r}
# Inspect possible values for SampleType
unique(tse$SampleType)
```

```{r eval = FALSE}
# Show the frequency of each value
tse$SampleType %>% table()
```

```{r echo = FALSE}
# Show the frequency of each value
tse$SampleType %>% table() %>% kable() %>%
    kableExtra::kable_styling("striped", latex_options="scale_down") %>%
    kableExtra::scroll_box(width = "100%")
```

::: {.callout-note}
## Note

After subsetting, expect the number of columns to equal the
sum of the frequencies of the samples that you are interested
in. For instance, `ncols = Feces + Skin + Tongue = 4 + 3 + 2 = 9`.
:::


Next, we _logical index_ across the columns of `tse` (make sure to
leave the first index empty to select all rows) and filter for the
samples of human origin. For this, we use the information on the
samples from the meta data `colData(tse)`.

```{r}
# Subset by sample
tse_sub <- tse[ , tse$SampleType %in% c("Feces", "Skin", "Tongue")]

# Show dimensions
dim(tse_sub)
```

As a sanity check, the new object `tse_subset_by_sample` should have
the original number of features (rows) and a number of samples
(columns) equal to the sum of the samples of interest (in this case
9).

Several criteria can be used to subset by sample:

* origin
* sampling time
* sequencing method
* DNA / RNA barcode
* cohort

### Subset by feature (row-wise)

Similarly, here we will extract a subset containing only the features
that belong to the phyla Actinobacteria and Chlamydiae, stored as
`Phylum` within `rowData(tse)`. However, subsetting by feature implies
a few more obstacles, such as the presence of `NA` elements and the
possible need for agglomeration.

As previously, we would first like to see all the possible values that `Phylum`
can have and how frequent those are:

```{r}
# Inspect possible values for phylum
# and show the first five values
cat(paste(unique(rowData(tse)$Phylum)[1:5], collapse = "\n"))

```

```{r}
# Show the frequency of each value
rowData(tse)$Phylum %>% table() %>% head() %>%
  kable() %>%
  kableExtra::kable_styling("striped", latex_options="scale_down") %>%
  kableExtra::scroll_box(width = "100%")

```

::: {.callout-note}
## Note

After subsetting, expect the number of columns to equal the
sum of the frequencies of the feature(s) that you are interested
in. For instance, `nrows = Actinobacteria + Chlamydiae = 1631 + 21 = 1652`.
:::

Depending on your research question, you might or might not need to
agglomerate the data in the first place: if you want to find the
abundance of each and every feature that belongs to Actinobacteria and
Chlamydiae, agglomeration is not needed. However, if you want to find the total
abundance of all features that belong to Actinobacteria or
Chlamydiae, agglomeration is recommended (see
[@sec-agglomeration] for details).

### Select features based on taxonomy rank

Next, we _logical index_ across the rows of `tse` (make sure to leave
the second index empty to select all columns) and filter for the
features that fall in either Actinobacteria or Chlamydiae group. For this,
we use the information on the samples from the metadata
`rowData(tse)`.

The first term with the `%in%` operator includes all the features
of interest, whereas the second term after the AND operator `&`
filters out all features that have an `NA` in place of the phylum variable.

```{r}
# Subset by feature
tse_sub <- tse[
    rowData(tse)$Phylum %in% c("Actinobacteria","Chlamydiae"), ]

# Show dimensions
dim(tse_sub)
```

As a sanity check, the new object, `tse_subset_by_feature`, should have the
original number of samples (columns) and a number of features (rows) equal to
the sum of the features of interest (in this case, 1652).

### Subset by sample and feature

Finally, we can subset data by sample and feature at once. The
resulting subset contains all the samples of human origin and all the
features of phyla Actinobacteria or Chlamydiae.

```{r}
# Subset by sample and feature and remove NAs
tse_sub <- tse[
  rowData(tse)$Phylum %in% c("Actinobacteria", "Chlamydiae") &
      !is.na(rowData(tse)$Phylum),
  tse$SampleType %in% c("Feces", "Skin", "Tongue")]

# Show dimensions
dim(tse_sub)
```

::: {.callout-note}
## Note

The dimensions of `tse_sub` are on par with
those of the previous subsets (9 columns filtered by sample and 1652
rows filtered by feature).
:::

If a study was to consider and quantify the presence of Actinobacteria
as well as Chlamydiae in different sites of the human body,
`tse_sub` might be a suitable subset to start
with.

### Remove empty columns and rows

Sometimes data might contain, e.g., features that are not present in any of
the samples. This can occur, for example, after data subsetting. In certain
analyses, we might want to remove those instances.

```{r}
# Agglomerate data at Genus level
tse_genus <- agglomerateByRank(tse, rank = "Genus", na.rm = FALSE)
# List bacteria that we want to include
genera <- c(
    "Class:Thermoprotei",
    "Genus:Sulfolobus",
    "Genus:Sediminicola")
# Subset data
tse_sub <- tse_genus[genera, ]

tse_sub
```

```{r}
# List total counts of each sample
colSums(assay(tse_sub, "counts"))[1:5]
```

Now we can see that certain samples do not include any bacteria. We can remove
those.

```{r}
# Remove samples that do not contain any bacteria
tse_genus_sub <- tse_sub[
  , colSums(assay(tse_sub, "counts")) != 0]
tse_sub
```

The same action can also be applied to the features.

```{r}
# Take only those samples that are collected from feces, skin, or tongue
tse_sub <- tse_genus[
    , tse_genus$SampleType %in% c("Feces", "Skin", "Tongue")]

tse_sub
```

```{r}
# What is the number of bacteria that are not present?
sum(rowSums(assay(tse_sub, "counts")) == 0)
```

We can see that there are bacteria that are not present in the samples we chose.
We can remove those bacteria from the data.

```{r}
# Take only those bacteria that are present
tse_sub <- tse_sub[
  rowSums(assay(tse_sub, "counts")) > 0, ]

tse_sub
```

## Splitting

You can split the data based on variables by using the functions 
`agglomerateByRanks()` and `splitOn()`.

`agglomerateByRanks()` splits the data based on taxonomic ranks. Since the 
elements of the output list share columns, they can be stored into `altExp` by
setting `as.list = FALSE`.

```{r agglomerateByRanks}
agglomerateByRanks(tse, as.list = TRUE)
```

If you want to split the data based on another variable than taxonomic rank, use 
`splitOn()`. It works for row-wise and column-wise splitting.

```{r splitOn}
splitOn(tse, "SampleType")
```

## Add or modify data {#sec-add-or-modify-data}

The information contained by the `colData` of a `TreeSE` can be modified by
accessing the desired variables.

```{r modify-coldata}
# modify the Description entries
colData(tse)$Description <- paste(
    colData(tse)$Description, "modified description")

# view modified variable
head(tse$Description)
```

New information can also be added to the experiment by creating a new variable.

```{r add-coldata}
# simulate new data
new_data <- runif(ncol(tse))

# store new data as new variable in colData
colData(tse)$NewVariable <- new_data

# view new variable
head(tse$NewVariable)
```

## Melting data

For several custom analysis and visualization packages, such as those from
`tidyverse`, the `SE` data can be converted to a long `data.frame` format with 
`meltSE()`.    

```{r}
library(mia)
data(GlobalPatterns, package="mia")
tse <- GlobalPatterns
tse <- transformAssay(tse, MARGIN = "cols", method="relabundance")
# Melt SE object
molten_tse <- mia::meltSE(
    tse,
    add.row = TRUE,
    add.col = TRUE,
    assay.type = "relabundance")
# Show the first five columns
molten_tse[, 1:5]
```

## Merge data

`mia` package has `mergeSEs()` function that merges multiple
`SummarizedExperiment` objects. For example, it is possible to combine
multiple `TreeSE` objects which each includes one sample.

`mergeSEs()` works much like standard joining operations. It combines rows and 
columns and allows you to specify the merging method.

```{r merge1}
# Take subsets for demonstration purposes
tse1 <- tse[, 1]
tse2 <- tse[, 2]
tse3 <- tse[, 3]
tse4 <- tse[1:100, 4]
```

```{r merge2}
# With inner join, we want to include all shared rows. When using mergeSEs
# function all samples are always preserved.
tse <- mergeSEs(list(tse1, tse2, tse3, tse4), join = "inner")
tse
```

```{r merge3}
# Left join preserves all rows of the 1st object
tse <- mergeSEs(tse1, tse4, missing.values = 0, join = "left")
tse
```
